---
title: "SNAP ALAAM analysis"
author: "Jonah Berman"
date: 5/26/2025
format: 
  pdf:
      output-file: "SNAP_ALAAM"
      output-ext: "pdf"
      toc: true
      toc-depth: 4
      shift-heading-level-by: 2
      fig-pos: "H"
      fig-cap-location: top
      geometry:
        - top=1in
        - right=.8in
        - bottom=1in
        - left=.8in
      link-citations: true
      linkcolor: blue
      include-in-header: 
        text: |
          \usepackage{fancyhdr}
          \usepackage{titling}
          \pagestyle{fancy}
          \fancyhf{}
          \renewcommand\maketitle{
            \fancyhead[C]{
              \thetitle
              \ifx \theauthor\empty  \else \ – \theauthor \fi
              \ifx \thedate\empty  \else \ – \thedate \ \fi
            }
          }
          \fancyfoot[C]{\thepage}
---

```{r}
#| label: libraries
#| echo: false
#| output: false
#| message: false

# Start with a clear environment
rm(list=ls())

# Load necessary packages
list.of.packages <- c("mvtnorm", "xtable","statnet","dplyr", "MASS", "coda", "devtools")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
library(devtools)
library(mvtnorm)
library(xtable)
library(statnet)
library(dplyr)
library(MASS)
library(coda)

# Load required libraries
library(tidyverse)
library(igraph)
library(statnet)

# Read in the Reddit hyperlink dataset
reddit_links <- read_tsv("soc-redditHyperlinks-body.tsv")

# Look at the structure
glimpse(reddit_links)



source("https://raw.githubusercontent.com/johankoskinen/ALAAM/main/MultivarALAAMalt.R") ## Bayesian ALAAM has not yet been released as an R library so we are using code from the github of the package creator, Johan Koskinen, Professor of Statistics at the University of Stockholm
# this will add lots of functions into your environment

# ------------------------------------------------------------------------------
# Set the working directory:
# Session > Set Working Directory > To Source File Location
# ------------------------------------------------------------------------------

list.files() 
library(igraph)
library(dplyr)
reddit_graph <- reddit_links %>%
  dplyr::select(SOURCE_SUBREDDIT, TARGET_SUBREDDIT) %>%
  filter(SOURCE_SUBREDDIT != TARGET_SUBREDDIT) %>%
  distinct() %>%
  graph_from_data_frame(directed = TRUE)


subreddits <- V(reddit_graph)$name
adj_reddit <- as.matrix(as_adjacency_matrix(reddit_graph, sparse = FALSE))
att <- data.frame(
  subreddit = subreddits,
  out_degree = degree(reddit_graph, mode = "out"),
  in_degree = degree(reddit_graph, mode = "in")
)

# filter by sentiment and take sample of 1500
sentiment_scores <- reddit_links %>%
  filter(!is.na(LINK_SENTIMENT)) %>%
  group_by(SOURCE_SUBREDDIT) %>%
  summarise(mean_sentiment = mean(LINK_SENTIMENT, na.rm = TRUE),
            post_count = n()) %>%
  filter(post_count >= 20)
echo_subs <- sentiment_scores %>%
  filter(mean_sentiment < 0) %>%
  slice_max(post_count, n = 500)
non_echo_subs <- sentiment_scores %>%
  filter(mean_sentiment >= 0) %>%
  slice_max(post_count, n = 1000)
stratified_sample <- bind_rows(echo_subs, non_echo_subs)
top_subs <- stratified_sample$SOURCE_SUBREDDIT

filtered_links <- reddit_links %>%
  filter(SOURCE_SUBREDDIT %in% top_subs,
         TARGET_SUBREDDIT %in% top_subs)

reddit_graph <- graph_from_data_frame(filtered_links, directed = TRUE)

att <- data.frame(
  subreddit = V(reddit_graph)$name,
  out_degree = degree(reddit_graph, mode = "out"),
  in_degree = degree(reddit_graph, mode = "in")
)

att <- left_join(att, sentiment_scores, by = c("subreddit" = "SOURCE_SUBREDDIT")) %>%
  mutate(echo_chamber = ifelse(mean_sentiment < 0, 1, 0)) %>%
  filter(!is.na(echo_chamber))

filtered_graph <- induced_subgraph(reddit_graph, vids = att$subreddit)
adj_reddit <- as.matrix(as_adjacency_matrix(filtered_graph, sparse = FALSE))

att$rec.ties <- rowSums(adj_reddit * t(adj_reddit))
att$triangles <- rowSums(adj_reddit * (adj_reddit %*% t(adj_reddit)))

DV <- att$echo_chamber

covs <- att %>%
  dplyr::select(out_degree, in_degree, rec.ties, triangles) %>%
  replace_na(list(out_degree = 0, in_degree = 0, rec.ties = 0, triangles = 0)) %>%
  as.matrix()
```

# SNAP Project ALAAM Analysis

The above code filters out a sample size of 1500 from our reddit network because the entire network was too cumbersome to run. Intentionally, we have take 500 of negative sentiment, which is this analysis is our dependent variable. And, we have taken 1000 with non-negative sentiment. The sampling was done to create better mixing. We will analyze which factors between in-degree, out-degree, contagion, triangles, and rec.ties are predictive or positive or negative sentiment. Echo chambers have been scientifically linked to negative sentiment. Thus, this analysis should find correlations between our factors and echo-chamber predictive negativity. 

#### **Hypothesis 1:** There is a negative effect on negative sentiment for in-degree

#### **Hypothesis 2:** There is a negative effect on negative sentiment for out-degree

#### **Hypothesis 3:** There is a positive effect on negative sentiment for contagion


::: {.callout-note style="color: purple"}
Please do NOT disseminate the code or data used in this assignment without permission.
:::

```{r}
#| label: run-alaam
#| cache: true
#| 
# test
res.0 <- BayesALAAM(
  y = DV,
  ADJ = adj_reddit,
  covariates = covs,
  directed = TRUE,
  Iterations = 1000,
  burnin = 100,
  contagion = c("simple"),
  saveFreq = 1000
)

Propsigma <- cov(res.0$Theta)
res.1 <- BayesALAAM(
  y = DV,
  ADJ = adj_reddit,
  covariates = covs,
  directed = TRUE,
  Iterations = 20000,
  burnin = 1000,
  contagion = c("simple"),
  saveFreq = 20000,
  PropSigma = Propsigma
)
```

```{r}
plot(ts(res.1$Thetas))
```

```{r}
write.res.table(
  burnin = 1,
  datamat = res.1$Thetas,
  thin = 1,
  tabname = "ALAAM_results"
)

temp <- read.csv("ALAAM_results.csv")
temp <- cbind(temp,
              ESS = res.1$ResTab[,"ESS"],
              SACF10 = res.1$ResTab[,"SACF 10"],
              SACF30 = res.1$ResTab[,"SACF 30"])
```




# PART I: Dataset


#### **Hypothesis 1:** Hypothesis 1 is supported.

We reject the null and accept this hypothesis. Greater in-degree increased negative sentiment by an odds ratio of 0.83. thus, each increase in in-degree results in negative-sentiment likelihood increasing by 17%

#### **Hypothesis 2:** Not statistically significant

P-value included 0. Results are not statistically significant despite showing negative correlation as hypothesis predicted

#### **Hypothesis 3:** Not able to confirm insights

Even though the CI doesn't include 0, this is highly uncertain effect and suggests unstable results. The plot lacks density and waves unpredictably. Likely there is structural sparsity or model sensitivity that prevents us from being able to use this extreme result.

### Overall key findings: 

Our analysis used ALAAM to analyze structural predictors of negative sentiment behavior across a 1,500 Reddit subreddit subset. We found that in-degree, which measures how often a subreddit is linked to by others, was a statistically significant negative predictor of negative sentiment. This suggests to us that more popular subreddits are less likely to exhibit negative sentiment. Thus, echo-chambers may be more likely to arise in smaller subreddits. In our study, other structural factors: out-degree, reciprocity, triadic closure, and contagion, were not significant or resulted in uncertain estimates. Our results tell us that negative sentiment formation on Reddit may be more influenced by node-level visibility and attention than by network diffusion or clustering via this ALAAM examination.




I used GPT and Copilot to assist with code fitting for this analysis Much of the code is repurposed from lab 5.

```{r}
#| label: save env
save.image(file = "SNAPALAAM.RData")
```

Please upload each file separately -- do not upload as a zip file! *(Please)*
